Reforecast analysis 4/6/2025
Maria Popescu

repository link: https://github.com/MariaPopescu123/MariaNEON-forecast-challenge-workshop

1.  Accessing climatology, persistence, and using MY MODEL for "BARC"

```{r}
library(tidyverse)
library(arrow)
library(kableExtra)
library(scoringRules)
library(duckdbfs)

targets <- read_csv("https://sdsc.osn.xsede.org/bio230014-bucket01/challenges/targets/project_id=neon4cast/duration=P1D/aquatics-targets.csv.gz")

```

Download all the forecasts generated by climatology, persistence, and MY MODEL.
I am going to do from January 2025-January 2026

    ```{r}
    all_forecasts_full <- duckdbfs::open_dataset(
      "s3://bio230014-bucket01/challenges/forecasts/bundled-parquet/project_id=neon4cast/duration=P1D/variable=temperature",
      s3_endpoint = "sdsc.osn.xsede.org",
      anonymous = TRUE
    ) |>
      filter(reference_datetime >= as_datetime("2025-01-01 00:00:00"),
             reference_datetime <= as_datetime("2026-01-01 00:00:00"),
         site_id == "BARC",
         model_id %in% c("climatology", "persistenceRW")) |> 
  collect() 
    
#to be able to plot performance over the course of horizon I need a column for horizon


    #write_csv(all_forecasts_full, "all_forecasts_full.csv")
    ```

9.  Calculating CRPS for climatology, persistence, and your model for all reference datetimes and horizons. Which model has the lower score when you consider all submissions?

    ```{r}
    obs_all <- targets |>
      mutate(datetime = as_datetime(datetime)) |>
      filter(variable == "temperature", site_id == "BARC") |>
      select(datetime, observation)

    # Climatology (normal)
    clim_all <- all_forecasts_full |>
      filter(model_id == "climatology") |>
      pivot_wider(names_from = parameter, values_from = prediction) |>
      inner_join(obs_all, by = "datetime") |>
      filter(!is.na(observation), !is.na(mu), !is.na(sigma)) |>
      mutate(crps = crps_norm(y = observation, mean = mu, sd = sigma)) |>
      select(reference_datetime, datetime, model_id, crps)

    # Persistence models
    ens_all <- all_forecasts_full |>
      filter(model_id %in% c("persistenceRW")) |>
      inner_join(obs_all, by = "datetime") |>
      filter(!is.na(observation), !is.na(prediction)) |>
      group_by(reference_datetime, datetime, model_id) |>
      summarise(crps = crps_sample(y = unique(observation), dat = prediction),
                .groups = "drop")
    ```

Now adding in my forecast
```{r}
#|include: false
library(tidyverse)
library(lubridate)
library(zoo)
library(neon4cast)

my_model_id <- 'maria_example_ID'
focal_sites <- c("BARC")
met_variables <- c("air_temperature")
forecast_horizon <- 35
n_members <- 31

# read in the targets data
targets <- read_csv("https://sdsc.osn.xsede.org/bio230014-bucket01/challenges/targets/project_id=neon4cast/duration=P1D/aquatics-targets.csv.gz")
 
targets <- targets|>
  filter(site_id %in% focal_sites,variable == 'temperature')

weather_past_s3 <- neon4cast::noaa_stage3()

weather_past <- weather_past_s3 |>
  dplyr::filter(site_id %in% focal_sites,
                datetime >= ymd('2017-01-01'),
                variable %in% met_variables) |>
  dplyr::collect()

weather_past_daily <- weather_past |>
  mutate(datetime = as_date(datetime)) |>
  group_by(datetime, site_id, variable) |>
  summarize(prediction = mean(prediction, na.rm = TRUE), .groups = "drop") |>
  mutate(prediction = ifelse(variable == "air_temperature", prediction - 273.15, prediction)) |>
  pivot_wider(names_from = variable, values_from = prediction)


all_forecast_dates <- seq(from = ymd("2025-01-01"), to = ymd("2025-12-31"), by = "week")

my_forecasts_full <- NULL

for (fdate in all_forecast_dates){
  forecast_date <- as_date(fdate)  
  noaa_date <- forecast_date - days(1)

  weather_future <- tryCatch({
    neon4cast::noaa_stage2(start_date = as.character(noaa_date)) |>
      dplyr::filter(datetime >= forecast_date,
                    site_id %in% focal_sites,
                    variable %in% met_variables) |>
      collect()
  }, error = function(e) { message("NOAA pull failed for ", forecast_date, ": ", e$message); NULL })

  if (is.null(weather_future) || nrow(weather_future) == 0) {
    message("skipping ", forecast_date, " (no NOAA data)")
    next
  }

weather_future_daily <- weather_future |>
  mutate(datetime = as_date(datetime)) |>
  # mean daily forecasts at each site per ensemble
  group_by(datetime, site_id, parameter, variable) |>
  summarize(prediction = mean(prediction, na.rm = TRUE), .groups = "drop") |>
  # convert air temperature to Celsius if it is included in the weather data
  mutate(prediction = ifelse(variable == "air_temperature", prediction - 273.15, prediction)) |>
  pivot_wider(names_from = variable, values_from = prediction) |>
  select(any_of(c('datetime', 'site_id', met_variables, 'parameter')))

#--------------------------#

forecasted_dates <- seq(from = forecast_date, to = forecast_date + forecast_horizon, by = "day")

# ----- Fit model & generate forecast----

#initial condition uncertainty (from water temp done)
#parameter uncertainty (done)
#driver uncertainty (from the 31 NOAA ensemble members)
#process uncertainty (done)


#only use targets from before reference datetime
# so the reforecast at each reference date doesn't peek at future observations
targets_lm <- targets |>
  pivot_wider(names_from = 'variable', values_from = 'observation') |>
  full_join(weather_past_daily,
            by = c("datetime","site_id"))|>
  filter(datetime < forecast_date) |>
  group_by(site_id)|>
  arrange(datetime)|>
  mutate(temp_yday = lag(temperature),
         air_temp_week_avg = lag(rollmean(air_temperature, k = 7, fill = NA, align = "right"), 1))|>
  ungroup()
#this is the average of the day before + 6 days prior


# Loop through each site to fit the model
forecast_df <- NULL

for(i in 1:length(focal_sites)) { 
  
    curr_site <- focal_sites[i] #UNCOMMENT WHEN DONE TESTING
  # curr_site <- "PRPO" #COMMENT WHEN DONE TESTING
  
  site_target <- targets_lm |>
    filter(site_id == curr_site)
  #THERE IS A GAP THOUGH. THE LAST DATE FOR AIR TEMP IS 4 WEEKS AGOOOOOO
  #no air temp past 2025
  
  noaa_future_site <- weather_future_daily |> 
    filter(site_id == curr_site)
  
  weather_ensemble_names <- unique(noaa_future_site$parameter)
  
  
  #Fit linear model based on past data: water temperature = b1 + b2 * yesterday's water temp + b3 * air temperature
  #you will need to change the variable on the left side of the ~ if you are forecasting oxygen or chla
  fit <- lm(site_target$temperature ~ site_target$temp_yday +
              site_target$air_temp_week_avg) #this is the avg fROM last 7 days
  
  #parameter uncertainty
  coeffs <- round(fit$coefficients, 2)
  fit_summary <- summary(fit)
  params_se <- fit_summary$coefficients[,2]
  mod <- predict(fit, site_target) #replace model_data with site_target
  
  #this part needed for parameter uncertainty
  param_df <- data.frame(beta1 = rnorm(n_members, coeffs[1], params_se[1]),
                         beta2 = rnorm(n_members, coeffs[2], params_se[2]),
                         beta3 = rnorm(n_members, coeffs[3], params_se[3]) #keeping this the same bc I removed air temp w no lag
                         ) 
  
  #this part needed for process uncertainty
  residuals <- mod - site_target$temperature
  sigma <- sd(residuals, na.rm = TRUE) # Process Uncertainty Noise Std Dev.; this is your sigma
  #used ai here^ to help me figure out how to introduce process uncertainty
  #but I understand that process uncertainty is what the model essentially doesn't capture. 
  #it's just the difference between model temperature and what the actual temperature is 
  
  #this part needed for IC uncertainty
  #need to update forecast_date to max datetime, because forecast date is not in the dataframe?
  curr_wt <- site_target %>% #need to update lake_df to site_target
    filter(!is.na(temperature))|>
    filter(datetime == max(datetime)) %>% # wtemp to temperature
    pull(temperature)
  
  ic_sd <- 0.1
  ic_uc <- rnorm(n = n_members, mean = curr_wt, sd = ic_sd) #this adds random noise based around my current wt to produce IC uncertainty
  
  #for each ensemble member there is now a diff IC value saved in this df to start it off
  #this from module 6 rmd
  ic_df <- tibble(forecast_date = rep(as.Date(forecast_date), times = n_members),
                  ensemble_member = c(1:n_members),
                  forecast_variable = "water_temperature",
                  value = ic_uc,
                  uc_type = "initial_conditions")
  
  forecast_ic_unc <- tibble(forecast_date = rep(forecasted_dates, times = n_members),
                            ensemble_member = rep(1:n_members, each = length(forecasted_dates)),
                            forecast_variable = "water_temperature",
                            value = as.double(NA),
                            uc_type = "initial_conditions") %>%
    rows_update(ic_df, by = c("forecast_date","ensemble_member","forecast_variable",
                              "uc_type")) 
  
  forecast_air_temp <- tibble (forecast_date = rep(forecasted_dates, times = n_members),
                               ensemble_member = rep(1:n_members, each = length(forecasted_dates)),
                               air_temperature = as.double(NA)) #this is empty at first
  
  #here will make similar frame to store as IC uncert for tracking lagged air temp
  
  # Loop through all forecast dates
  for (t in 1:length(forecasted_dates)) {
    
     #t <- 4 #COMMENT WHEN DONE TESTING
    
    repeat_ens <- rep(weather_ensemble_names, length.out = n_members) #this is 0-30 repeated 
    
    # Loop over each ensemble member
    for(ens in 1:n_members){ 

    
      # ens <- 2 #COMMENT WHEN DONE TESTING
      
      met_ens <- repeat_ens[ens] #now I am grabbing from the index in repeat_ens 
      
      
      #pull driver ensemble for the relevant date; here we are using all 310 NOAA ensemble members
      temp_driv <- weather_future_daily %>%
        filter(datetime == forecasted_dates[t],
               site_id == curr_site,
               parameter == met_ens)
      #this is just daily though, need to compute avg prev 7 days now
      #need to forecast air temp so that we can calculate
      # forecast_air_temp <- forecast_air_temp|>
      #   mutate(air_temperature = ifelse(
      #     forecast_date == forecasted_dates[t] & ensemble_member == ens, 
      #          temp_driv$air_temperature[1], 
      #     air_temperature
      #     ))
      
      #now need to combine forecasted air temp with the historical data 
      #going to select (current date - 7) as a filter basically then average 
      #BUT NOW REALIZING BC OF THE GAP THIS WILL NEVER WORK. 
      #MAYBE I INSTEAD THE FIRST AVG_WEEKLY WILL JUST BE WHAT THE LAST AVAILABLE ONE WAS 
      #THIS WILL BE INNACURATE BUT A SOLUTION
      
      #df w 7 prior dates 
      forecasted_air_seven <- forecast_air_temp|> 
        filter(ensemble_member == ens, 
               forecast_date < forecasted_dates[t], #the forecast dates before this current one
               forecast_date >= forecasted_dates[t] - days(7))|> #<- help w AI here. didn't know I could use days(7)
        select(forecast_date, air_temperature) |>
        rename(datetime = forecast_date)
      
      #this actually probs not necessary anymore
      #now just bind to historical everything and then filter for the last 7 days
      #site_target is our target 
      site_target_seven <- site_target|>
        filter(datetime >= forecasted_dates[t] - days(7), 
               datetime < forecasted_dates[t])|>
        select(datetime, air_temperature)
      
      combined_seven <- bind_rows(forecasted_air_seven, site_target_seven)|>
        filter(datetime >= forecasted_dates[t] - days(7), 
               datetime < forecasted_dates[t])
      
      if(t == 1){
        air_temp_week_avg <- site_target |>
          filter(datetime == max(datetime, na.rm = TRUE)) |> #I know this is weird, because it doesn't make sense that with the time gap, the value would be the same, but this is my temporary solution so that I at least have a forecast
          pull(air_temp_week_avg) 
      }else {air_temp_week_avg <- mean(combined_seven$air_temperature, na.rm = TRUE)
}
        
      # pull lagged water temp: use IC uncertainty for first date, previous forecast for subsequent dates
      #had difficulty in this part with the for loop and integrating ensemble members, had help w AI
      
      if(t == 1){
        temp_lag <- forecast_ic_unc %>%
          filter(forecast_date == forecasted_dates[t],
                 ensemble_member == ens) %>%
          pull(value)
      } else {
        temp_lag <- forecast_ic_unc %>%
          filter(forecast_date == forecasted_dates[t-1],
                 ensemble_member == ens) %>%
          pull(value)
      }
      
      
      
      #updated for param + IC + process uncertainty: wt = b1 + [b2 * yesterday's wt] + [this part replaced by air temp avg] + W
      forecasted_temperature <- param_df$beta1[ens] + param_df$beta2[ens] * temp_lag + param_df$beta3[ens] * air_temp_week_avg + rnorm(1, mean = 0, sd = sigma) #<- ai helped me integrate process uncertainty here 
      #param uncertainty coming from the coefficients calculated earlier
      #IC uncertainty is set up before the for loop and integrated in the first time step when I make the dataframe
      #process uncertainty is the last part of this equation calculated from sigma earlier from the difference between the model and the observed temperatures. 
      
      # store forecast back into forecast_ic_unc so it can be used as lag for next timestep
      forecast_ic_unc <- forecast_ic_unc %>%
        mutate(value = ifelse(forecast_date == forecasted_dates[t] & ensemble_member == ens,
                              forecasted_temperature, value))
      
      #store air temp same way storing ylag for wtemp
      forecast_air_temp <- forecast_air_temp %>%
        mutate(air_temperature = ifelse(
          forecast_date == forecasted_dates[t] & ensemble_member == ens,
          temp_driv$air_temperature[1],
          air_temperature
        ))
      
      #before forecasting computing average for last 7 days (bc this might include forecasted dates 
      #AND historical dates)
      
      # put all the relevant information into a tibble that we can bind together
      curr_site_df <- tibble(datetime = forecasted_dates[t],
                             site_id = curr_site,
                             parameter = met_ens,
                             prediction = forecasted_temperature,
                             variable = "temperature") #Change this if you are forecasting a different variable
      
      
      forecast_df <- dplyr::bind_rows(forecast_df, curr_site_df)
      
    }
  }
  
  message(curr_site, ' forecast run')
  
}
 
  # ---- Format to EFI standard and accumulate ----
  if (!is.null(forecast_df) && nrow(forecast_df) > 0) {
    forecast_df_EFI <- forecast_df |>
      filter(datetime > forecast_date) |>
      mutate(
        model_id           = my_model_id,
        reference_datetime = as_datetime(forecast_date),
        datetime           = as_datetime(datetime),
        family             = 'ensemble',
        duration           = 'P1D',
        parameter          = as.character(parameter),
        project_id         = 'neon4cast'
      ) |>
      select(datetime, reference_datetime, duration, site_id,
             family, parameter, variable, prediction, model_id, project_id)

    my_forecasts_full <- dplyr::bind_rows(my_forecasts_full, forecast_df_EFI)
  }

  message("Completed: ", forecast_date)

}

    mine_all <- my_forecasts_full |>
      inner_join(obs_all, by = "datetime") |>
      filter(!is.na(observation), !is.na(prediction)) |>
      group_by(reference_datetime, datetime, model_id) |>
      summarise(crps = crps_sample(y = unique(observation), dat = prediction),
                .groups = "drop")

```

```{r}
#| include: false
    crps_full <- bind_rows(clim_all, ens_all, mine_all)|>
        mutate(horizon = as.numeric(difftime(datetime, reference_datetime, units = "days"))) #for the horizon

    mean_crps_by_horizon <- crps_full |>
      group_by(model_id, horizon) |>
      summarise(mean_crps = mean(crps, na.rm = TRUE),
                n = n(),
                .groups = "drop") |>
      arrange(mean_crps)
    mean_crps_by_horizon
```


```{r}
ggplot(mean_crps_by_horizon, aes(x = horizon, y = mean_crps, color = model_id))+
  geom_line() +
  geom_point() +
  labs(x = "horizon", y = "CRPS", color = "model")

```

So mine does better for the first 2 days then quickly gets worse than both models.  
